---
title: "Causal Survival Random Forests"
author: "Quinn Hungerford and Isabella Summe"
date: "2025-06-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading in libraries

```{r}
library(grf)
library(ggplot2)
library(survival)
library(survcomp)
library(pec)
library(tibble)
library(dplyr)
```

# Importing data, cleaning, and setting new vars

```{r}

# OLD IMPORTING + CLEANING (LEO's DATASET)

# Import data
data <- read.csv("rhc_imputed.csv")

# Keep only rows with complete outcome information
data_clean <- data[!is.na(data$SurvivalTime) & !is.na(data$death), ]

# Define treatment and survival outcome
W <- as.numeric(data_clean$swang1)
Y <- data_clean$SurvivalTime
D <- data_clean$death

# Define covariates
exclude <- c("ptid", "swang1", "dthdte", "dschdte", "lstctdte", "sadmdte",
             "survtime", "death", "SurvivalTime", "censor")
X_temp <- data_clean[, !(names(data_clean) %in% exclude)]

# Factor key categorical variables
X_temp$sex <- as.factor(X_temp$sex)
X_temp$race <- as.factor(X_temp$race)
X_temp$income <- as.factor(X_temp$income)
X_temp$cat1 <- as.factor(X_temp$cat1)
X_temp$ca <- as.factor(X_temp$ca)

# Convert to dummy variables
X <- model.matrix(~ . - 1, data = X_temp)
```

```{r}

# Import data
data <- read.csv("rhc_imputed.csv")
#data <- read.csv("new_data.csv")


# Adjusting variables

# Recalculate survival time

data$survtime_new <- ifelse(!is.na(data$dthdte),  # Check if they have a date of death
                            data$dthdte - data$sadmdte, # If true (they died), get time from admission to death
                            data$lstctdte - data$sadmdte) # If false (didn't), get time from admission to the last known contact date

# Recalculate event indicator -- if true/dead, assign 1 and if false/censored, assign 0

data$death_new <- ifelse(!is.na(data$dthdte), 1, 0)  # Check if they have a non-missing date of death

# Clean data so no NAs
data_clean <- data[!is.na(data$survtime_new) & !is.na(data$death_new), ]

# Treatment indicator W (RHC 1 or no RHC 0)
W <- as.numeric(data_clean$swang1)

# Event time Y (survival time) and event type D (death)
Y <- data_clean$survtime_new
D <- data_clean$death_new

# Non-covariates

exclude <- c("ptid", "swang1", "dthdte", "dschdte", "lstctdte", "sadmdte",
              "survtime", "death", "survtime_new", "death_new")

# Create covariate subset

X_temp <- data_clean[, !(names(data_clean) %in% exclude)]

# Make sure important covariates are factored

X_temp$sex <- as.factor(X_temp$sex)
X_temp$race <- as.factor(X_temp$race)
X_temp$income <- as.factor(X_temp$income)
X_temp$cat1 <- as.factor(X_temp$cat1)
X_temp$ca <- as.factor(X_temp$ca)

# Convert all variables to dummy variables for the model
cc_idx <- complete.cases(X_temp)
X <- model.matrix(~ . - 1, data = X_temp[cc_idx, ])
Y <- Y[cc_idx]
D <- D[cc_idx]
W <- W[cc_idx]

```



# Checking for Truncation Time

```{r}
hist(Y[D == 1], main = "Event vs. Censored", xlab = "", ylim = c(0, 100), xlim = c(0, 200))
hist(Y[D == 0], col = adjustcolor("red", 0.5), add = TRUE)
legend("topright", c("Event", "Censored"),
col = c("gray", adjustcolor("red", 0.5)),
lwd = 4)
abline(v = 720, lty = 2)

```


# CSF Models - Survival Probability: 30 days, 60 days, 90 days, 120 days, 150 days, 180 days

```{r}

# failure_grid <- seq(0, max(Y), by = 5)  # every 5 days

csf_30_sb <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 30,
  num.trees = 1000,
  target = "survival.probability",
  #failure.times = failure_grid,
  seed = 2025
)

csf_60_sb <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 60,
  num.trees = 1000,
  target = "survival.probability",
  #failure.times = failure_grid,
  seed = 2025
)

csf_90_sb <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 90,
  num.trees = 1000,
  target = "survival.probability",
  #failure.times = failure_grid,
  seed = 2025
)

csf_120_sb <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 120,
  num.trees = 1000,
  target = "survival.probability",
  #failure.times = failure_grid,
  seed = 2025
)

csf_150_sb <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 150,
  num.trees = 1000,
  target = "survival.probability",
  #failure.times = failure_grid,
  seed = 2025
)

csf_180_sb <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 180,
  num.trees = 1000,
  target = "survival.probability",
  #failure.times = failure_grid,
  seed = 2025
)
```


# CSF Models -- RMST: 30 days, 60 days, 90 days, 120 days, 150 days, 180 days

```{r}
csf_30_rmst <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 30,
  num.trees = 1000,
  target = "RMST",
  #failure.times = failure_grid,
  seed = 2025
)

csf_60_rmst <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 60,
  num.trees = 1000,
  target = "RMST",
  #failure.times = failure_grid,
  seed = 2025
)

csf_90_rmst <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 90,
  num.trees = 1000,
  target = "RMST",
  #failure.times = failure_grid,
  seed = 2025
)

csf_120_rmst <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 120,
  num.trees = 1000,
  target = "RMST",
  #failure.times = failure_grid,
  seed = 2025
)

csf_150_rmst <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 150,
  num.trees = 1000,
  target = "RMST",
  #failure.times = failure_grid,
  seed = 2025
)

csf_180_rmst <- causal_survival_forest(
  X = X,
  Y = Y,
  D = D,
  W = W,
  horizon = 180,
  num.trees = 1000,
  target = "RMST",
  #failure.times = failure_grid,
  seed = 2025
)
```


# CATEs

```{r}

# Helper function to extract summary stats from a prediction vector
get_summary <- function(preds, horizon, target_type) {
  s <- summary(preds)
  tibble(
    Horizon = horizon,
    Target = target_type,
    Min = s["Min."],
    Q1 = s["1st Qu."],
    Median = s["Median"],
    Mean = s["Mean"],
    Q3 = s["3rd Qu."],
    Max = s["Max."]
  )
}

# Predict and summarize CATEs for survival probability
cate_summary_sp <- bind_rows(
  get_summary(predict(csf_30_sb)$predictions, 30, "Survival Probability"),
  get_summary(predict(csf_60_sb)$predictions, 60, "Survival Probability"),
  get_summary(predict(csf_90_sb)$predictions, 90, "Survival Probability"),
  get_summary(predict(csf_120_sb)$predictions, 120, "Survival Probability"),
  get_summary(predict(csf_150_sb)$predictions, 150, "Survival Probability"),
  get_summary(predict(csf_180_sb)$predictions, 180, "Survival Probability")
)

# Predict and summarize CATEs for RMST
cate_summary_rmst <- bind_rows(
  get_summary(predict(csf_30_rmst)$predictions, 30, "RMST"),
  get_summary(predict(csf_60_rmst)$predictions, 60, "RMST"),
  get_summary(predict(csf_90_rmst)$predictions, 90, "RMST"),
  get_summary(predict(csf_120_rmst)$predictions, 120, "RMST"),
  get_summary(predict(csf_150_rmst)$predictions, 150, "RMST"),
  get_summary(predict(csf_180_rmst)$predictions, 180, "RMST")
)

# Combine for final table
#cate_summary_all <- bind_rows(cate_summary_sp, cate_summary_rmst)

# View table
#print(cate_summary_all)

```

```{r}
cate_summary_sp
```

```{r}
cate_summary_rmst
```


```{r}
# Predict all CATEs
cate_estimates_sp <- list(
  `30` = predict(csf_30_sb)$predictions,
  `60` = predict(csf_60_sb)$predictions,
  `90` = predict(csf_90_sb)$predictions,
  `120` = predict(csf_120_sb)$predictions,
  `150` = predict(csf_150_sb)$predictions,
  `180` = predict(csf_180_sb)$predictions
)

cate_estimates_rmst <- list(
  `30` = predict(csf_30_rmst)$predictions,
  `60` = predict(csf_60_rmst)$predictions,
  `90` = predict(csf_90_rmst)$predictions,
  `120` = predict(csf_120_rmst)$predictions,
  `150` = predict(csf_150_rmst)$predictions,
  `180` = predict(csf_180_rmst)$predictions
)

```

```{r}
# Plot histograms for Survival Probabilty

par(mfrow = c(3, 2))  # 6 plots per batch

# Survival probability
for (h in names(cate_estimates_sp)) {
  hist(cate_estimates_sp[[h]], breaks = 50,
       main = paste("CATEs: Survival Prob. (", h, "-Day)", sep = ""),
       xlab = paste("Effect of RHC on", h, "Day Survival Probability"),
       col = "lightblue", border = "white")
}

```

```{r}
par(mfrow = c(3, 2))

# RMST
for (h in names(cate_estimates_rmst)) {
  hist(cate_estimates_rmst[[h]], breaks = 50,
       main = paste("CATEs: RMST (", h, "-Day)", sep = ""),
       xlab = paste("Effect of RHC on", h, "Day RMST"),
       col = "lightgreen", border = "white")
}
```



# Cross-Validation for Survival Probability Model
```{r}

set.seed(2025)
folds <- sample(1:5, size = length(Y), replace = TRUE)

# Horizon values
horizons <- c(30, 60, 90, 120, 150, 180)
c_index_sp <- numeric(length(horizons))

for (j in seq_along(horizons)) {
  h <- horizons[j]
  c_indices <- numeric(5)
  
  for (i in 1:5) {
    train_idx <- which(folds != i)
    test_idx  <- which(folds == i)

    # Train model
    model_cv <- causal_survival_forest(
      X = X[train_idx, ],
      Y = Y[train_idx],
      D = D[train_idx],
      W = W[train_idx],
      horizon = h,
      target = "survival.probability",
      num.trees = 1000,
      seed = 2025
    )

    # Predict survival probability and convert to risk scores
    preds <- predict(model_cv, newdata = X[test_idx, ], failure.time = h)
    risk_scores <- 1 - preds$predictions
    surv_obj <- Surv(Y[test_idx], D[test_idx])

    # Calculate C-index
    concordance_result <- concordance(surv_obj ~ risk_scores)
    c_indices[i] <- concordance_result$concordance
  }

  c_index_sp[j] <- mean(c_indices)
}

# View results
c_index_sp_table <- tibble(
  Horizon = horizons,
  Target = "Survival Probability",
  Mean_C_Index = round(c_index_sp, 4)
)

print(c_index_sp_table)

```


# Cross-Validation for RMST
```{r}
set.seed(2025)
folds <- sample(1:5, size = length(Y), replace = TRUE)

c_index_rmst <- numeric(length(horizons))

for (j in seq_along(horizons)) {
  h <- horizons[j]
  c_indices <- numeric(5)
  
  for (i in 1:5) {
    train_idx <- which(folds != i)
    test_idx  <- which(folds == i)

    # Train model
    model_cv <- causal_survival_forest(
      X = X[train_idx, ],
      Y = Y[train_idx],
      D = D[train_idx],
      W = W[train_idx],
      horizon = h,
      target = "RMST",
      num.trees = 1000,
      seed = 2025
    )

    # Predict RMST and convert to risk scores (negative = worse outcome)
    preds <- predict(model_cv, newdata = X[test_idx, ], failure.time = h)
    risk_scores <- -preds$predictions  # Higher RMST = better, so reverse sign
    surv_obj <- Surv(Y[test_idx], D[test_idx])

    # Calculate C-index
    concordance_result <- concordance(surv_obj ~ risk_scores)
    c_indices[i] <- concordance_result$concordance
  }

  c_index_rmst[j] <- mean(c_indices)
}

# View results
c_index_rmst_table <- tibble(
  Horizon = horizons,
  Target = "RMST",
  Mean_C_Index = round(c_index_rmst, 4)
)

print(c_index_rmst_table)

```


```{r}
# View results
c_index_rmst_table <- tibble(
  Horizon = horizons,
  Target = "RMST",
  Mean_C_Index = round(c_index_rmst, 4)
)

print(c_index_rmst_table)
```


```{r}

#cate_estimates_sp
#cate_estimates_rmst



```



```{r}



# Calculates quantiles of the CATE estimates
quantile(cate_estimates, probs = c(0.01, 0.25, 0.5, 0.75, 0.99))

# Add CATE estimates to the data frame so each patient's effect is stored
data_clean$CATE <- cate_estimates

# Groups patients into categories based on their estimated treatment effect (CATE)
data_clean$Effect_Group <- cut(data_clean$CATE,
                                breaks = c(-Inf, -0.05, 0.05, Inf),
                                labels = c("Harmed", "Neutral",
                                           "Benefitted"))
# Harmed if RHC reduces survival by more than 0.05 days
# Neutral if the effect is between +-0.05
# Benefitted if RHC increases survival by more than 0.05 days

# Count how many patients are into each effect group
table(data_clean$Effect_Group)
```
The model predicts that most patients (about 83%) are harmed by RHC in terms of 30-day survival. A small subset (about 13%) may benefit, and only around 2% appear unaffected. This supports the presence of treatment effect heterogeneity + justifies us investigating which subgroups these benefitted individuals belong to.

# ATE (for reference)

```{r}
# Average Treatment Effect
ate <- mean(csf_preds$predictions)
ate
```

```{r}
average_treatment_effect(model)
```


# Identifying variables that drive the heterogeneity in how RHC affects survival (aka most contribute to variation in CATEs)

```{r}
# Key Drivers of HTE
vi <- variable_importance(model)
top_vars <- data.frame(
  Variable = colnames(X),
  Importance = vi
)
top_vars <- top_vars[order(-top_vars$Importance), ]
head(top_vars, 40)
```

# Graphic for the top 10 covariates

```{r}
# Top 10 vars
top10 <- head(top_vars, 10)

# Order the vars
top10$Variable <- factor(top10$Variable, levels = rev(top10$Variable))

# Plot
ggplot(top10, aes(x = Variable, y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Variables",
       x = "Variable",
       y = "Importance") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 12)
  )
```


# Bella's stuff that I do not understand lol


# dementhx, sex1, incomeUnder..11k
# t3d30, wtkilo1, neuro, seps, income.11..25k, income.25..50k

```{r}

best_linear_projection(csf_30_rmst, X)

```



# TOC

```{r}
tau.hat = predict(csf_150_sb)$predictions
# Form a doubly robust estimate of RATE using cs.forest
rate = rank_average_treatment_effect(csf_150_sb, tau.hat)
# Plot the TOC, along with 95 % confidence bars
plot(rate,
main = "TOC: By decreasing CATE estimates"
)
```

```{r}
tau.hat = predict(csf_30_rmst)$predictions
# Form a doubly robust estimate of RATE using cs.forest
rate = rank_average_treatment_effect(csf_30_rmst, tau.hat)
# Plot the TOC, along with 95 % confidence bars
plot(rate,
main = "TOC: By decreasing CATE estimates"
)
```
```{r}
paste("AUTOC:", round(rate$estimate, 2), "+/-", round(1.96 * rate$std.err, 2))
```

```{r}
q80 = quantile(tau.hat, 0.8)
rbind(
full.sample = apply(X, 2, mean),
top.20 = apply(X[tau.hat >= q80, ], 2, mean)
)
```


```{r}
# Get predicted treatment effects (tau-hat) for each individual
tau_hat <- predict(model)$predictions

# Run RATE with tau_hat as the ranking priority
rate_result <- rank_average_treatment_effect(model, priorities = tau_hat)

# Plot
plot(rate_result)
```

# SGATE

```{r}
# Get predicted treatment effects
tau_hat <- predict(csf_30_rmst)$predictions

# Split into quantile groups (e.g., quintiles)
groups <- cut(tau_hat, breaks = quantile(tau_hat, probs = seq(0, 1, length.out = 6)),
              include.lowest = TRUE, labels = FALSE)

# Estimate ATE within each group
library(dplyr)
sgate_df <- data.frame(tau_hat = tau_hat, group = groups)

sgate_df |> 
  group_by(group) |> 
  summarise(mean_tau = mean(tau_hat)) |> 
  ggplot(aes(x = factor(group), y = mean_tau)) +
  geom_col(fill = "steelblue") +
  labs(title = "S-GATE (Quintiles)", x = "Quantile Group", y = "Average Treatment Effect") +
  theme_minimal()
```


